MetaTracts - A Method for Robust Extraction and Visualization of Carbon Fiber Bundles in Fiber Reinforced Composites

### 2014-12-10 chapters/MetaTracts.tex
shortened the paper to fit within 8 pages

We demarcate the above queries into two parts: \textit{geometric structure} and \textit{spatial context}. Geometric structure refers to the shape, size, and orientation of a fiber bundle or a group of \replacedWith{bundles}{them}. Spatial context refers to how two or more bundles interact with each other. \replacedWith{For example, spatial context}{It provides} answers \added{to} questions such as: Are these bundles in contact at a particular position in the data? What is the relative orientations of the contacting bundles? 
\replacedWith{Answering}{Providing answers to} the above queries from the volume renderings of XCT datasets or from visual inspection of particular 2D slices is non-trivial, even for experts. We \replacedWith{detail}{present} and evaluate a method which uses visualization techniques to gain insight into our data.
We \replacedWith{advance and interpret}{interpret and advance} techniques from diffusion tensor imaging to extract and visualize geometric structures from \replacedWith{3D X-ray computed tomography}{XCT} data of the woven carbon fiber reinforced composites. The main goal of this work is to expand the state of the art in non-destructive testing through visualization of composite structures in \replacedWith{lower resolutions and thus larger areas}{\textit{complete unit cells}} of woven fabrics.
\begin{figure}[tb]
	\centering
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/algo.png}
	%\includegraphics[width=0.5\textwidth,  trim = 0mm 300mm 0mm 0mm, clip,]{imagesMT2014/image_algo}
	\includegraphics[width=\linewidth]{images_pvis/workflow.pdf}
	\caption{Flow chart of the MetaTracts approach for fiber bundle extraction}
	\label{fig:flowchart}
\end{figure}
Figure~\ref{fig:flowchart} shows an over view of our approach.
\replacedWith
{
Starting from the XCT data, we first generate MetaTracts, cluster the MetaTracts to create fiber bundles and then generate visualizations from the clusters. For a detailed description of the MetaTracts generation process, see Sec.~\ref{subsec:MetaTracts-generation}. For the description of plausible fiber bundle generation see Sec.~\ref{subsec:fiber-bundles}. For a discussion of the visualizations we create for our domain experts see Sec.~\ref{sec:vis}. Sec.~\ref{sec:results} describes our experimental results, Sec.~\ref{sec:param_choices} presents our parameter choices and  Sec.~\ref{sec:user_eval} summarizes our user evaluations.
} 
{
Starting from XCT data, we first generate MetaTracts (see Sec.~\ref{sec:approach}). Then we cluster the MetaTracts to generate and visualize fiber bundles (see Sec.~\ref{subsec:fiber-bundles}). We discuss the visualizations we created for our domain experts in Sec.~\ref{sec:vis}. Sec.~\ref{sec:results} describes our experimental results, Sec.~\ref{sec:param_choices} presents our parameter choices and  Sec.~\ref{sec:user_eval} finally summarizes the user evaluation of our method.
}

@@

\begin{figure}[tb]
\centering
%\includegraphics[width=0.5\textwidth, trim = 0mm 110mm 0mm 0mm, clip,]{images_pvis/figure1}
\includegraphics[width=\linewidth]{images_pvis/data-char.pdf}
\caption
{
\replacedWith
{
Data characteristics: (a) Rendering of the data. (b) 2D slices along the X-axis and (c) 2D slice along the Z-axis. (d) Magnified region marked in green square (c). Multiple fiber bundles cross and are indistinguishable by visual inspection alone. 
}
{
Data characteristics: (a) Rendering of dataset D1. (b, c) 2D slices along Z- and X-axis. (d) Zoom in of the green region marked in (c). Multiple fiber bundles cross and are indistinguishable by visual inspection alone. 
}
}
\label{fig:data-char}
\end{figure}

\section {Data Characteristic and Assumptions}
\label{sec:char_data}
\replacedWith
{
Figure \ref{fig:data-char} shows dataset 1 with woven fiber bundles. Dataset 1 is 450$\times$300$\times$500 voxels in size with isotropic resolution of 2$\mu m$. The data type is unsigned 8 bit integer. Figure~\ref{fig:data-char} clearly shows the weaving structure of the composite unit cell as regional subset of the recurring fiber bundle weaving pattern. The unit cell is used as a basis structure for CFRP manufacturing. The alignment and weaving pattern of the fiber bundles in the cloth influences the strength and stiffness properties of the resulting material. Figure~\ref{fig:data-char}a shows a volume rendering of the dataset. Figure~\ref{fig:data-char}b shows 2D slices along the $X$- and Figure~\ref{fig:data-char}c in $Z$-axis. Figure~\ref{fig:data-char}d shows the magnified version of the green region of interest. 
The resolution of dataset 1 renders the individual carbon fibers in a fiber bundle hard to resolve in the 2D slice images. Figure~\ref{fig:data-char}d contains two bundles going in opposite directions. The separation between two fiber bundles is barely visible. The carbon fibers themselves do not differ much from the underlying epoxy matrix in terms of attenuation which poses an additional challenge for characterization. The fiber bundles may differ in terms of the amount of fibers in the bundle. Figure~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern, the fiber bundles cross each other in different orientations. In general, the number of orientations is limited by the weaving pattern. Nevertheless, the weaving pattern may cause the individual bundles to be curved. In consequence individual fibers may be adjacent in Euclidean distance but belong to different bundles. 
We make the following assumptions on our data:
}
{
Figure~\ref{fig:data-char} shows dataset D1 with woven fiber bundles. The size of D1 is 450$\times$300$\times$500 voxels with isotropic resolution of 2 $\mu m$ and 8 bit unsigned integer scalars. Figure~\ref{fig:data-char} clearly shows the recurring fiber bundle weaving pattern of the composite unit cell used for manufacturing. Figure~\ref{fig:data-char}a shows a volume rendering of the dataset. Figure~\ref{fig:data-char}b and ~\ref{fig:data-char}c show 2D slices along the $X$- and the $Z$-axis respectively. Figure~\ref{fig:data-char}d zooms into the green region of interest. 
The resolution of D1 renders the individual carbon fibers in a fiber bundle hard to resolve in the 2D slice images. Figure~\ref{fig:data-char}d contains two bundles going in opposite directions. Also the separation between two fiber bundles is barely visible. The fiber bundles may differ in terms of the amount of fibers in the bundle. Figure~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern, the fiber bundles cross each other in different orientations. The number of orientations is defined by the weaving pattern and typically consists of two main orientations. The weaving pattern may cause individual bundles to be curved. In consequence, individual fibers may be adjacent in Euclidean distance but belong to different bundles. 
We make the following \textit{assumptions} on our data:
}

@@

\begin{figure}[tb]
\centering
%\includegraphics[width=0.45\textwidth]{images_pvis/figure3}
%\includegraphics[width=0.5\textwidth,  trim = 0mm 90mm 0mm 0mm, clip]{images_pvis/figure3}
\includegraphics[width=\linewidth]{images_pvis/reliable_hessian.pdf}
 \vspace{-1.5em}
\caption{Reliable Hessians. (a) Colored according to Orientation vector mapped to RGB. (b\added{,c}) 2D slice\added{s} along Z\replacedWith{ and (c) along}{-} and X axis. (d) \replacedWith{Magnified}{Zoom in of the green} region marked in \replacedWith{green square}{(c)}.}
\label{fig:reliable_hessian}
\end{figure}

@@

\subsection {MetaTracts generation}
\label{subsec:MetaTracts-generation}
In this section we explain the process of MetaTracts generation in $\mathbb{R}^2$ (the procedure trivially extends to $\mathbb{R}^3$). In $\mathbb{R}^2$, all the above properties hold except that cylinders are replaced by rectangles. The reliable Hessian step generates a local orientation vector in $\mathbb{R}^2$ and a reliability measure $R_H$[0,1]. In Figure~\ref{fig:algo}, regions with unreliable Hessian have been marked in blue. Figure~\ref{fig:algo}c shows \replacedWith{a single}{an individual} cylinder of a MetaTract. Let the seed point associated with the MetaTract be grid point $C_{p}$ (property 2 in Sec.~\ref{subsec:mtprop}). The local orientation at $C_p$ as computed in the step above is $N_P$ and is given by the dark green arrow. We generate a rectangle of length $L$ and radius $R$ (properties 3 and 5 in Section~\ref{subsec:mtprop}). 
%
\begin{figure}[tb]
  \begin{center}
 \includegraphics[ width=\linewidth]{images_pvis/mt_gen.pdf}
  \end{center}
  \vspace{-1em}
 \caption{MetaTracts generation: (b) \replacedWith{MetaTract generation, region with $R_H$ less than threshold}{Generation process}. (a) \replacedWith{Close up of}{Zoom in of the white} marked region in (b). (c) \replacedWith{Parts of a single MetaTract}{Individual cylinder}.}
  \label{fig:algo}
    \vspace{-1.5em}
\end{figure}
The set of vertices which are in the green region but not overlapped by blue are possible \textit{``candidate vertices"} for the start point of the next cylinder in a particular MetaTract. From these we select another grid vertex which will be the start point for the next cylinder.
%\begin{figure}
%\centering
% \includegraphics[width=0.25\textwidth]{imagesMetaTracts/angle-diff.png}
%\caption{Angle between the local orientation of two cylinders that are part of the same MetaTract.}
%\label{fig:angle-diff}
%\end{figure}
Based on assumptions on our data (local orientation and connectivity), the order of the candidate vertices is based on the following \textit{characteristics}:

@@

\begin{figure}[tb]
\centering
	\includegraphics[width=\linewidth]{images_pvis/figure5.pdf}
	%\includegraphics[width=0.45\textwidth,  trim = 0mm 100mm 0mm 50mm, clip]{imagesMT2014/image_length}
	\caption{(a) MetaTracts color-coded according to their mean orientations. (b) length distribution of individual MetaTracts for a particular bundle (unit for length is grid cube edge length:  $2\mu m$).}
	\label{fig:length_distribution}
\end{figure}

@@

\begin{figure}[tb] 
  \centering  	
  \includegraphics[width=\linewidth]{images_pvis/clustering.pdf}
   \vspace{-1.5em}
  %\includegraphics[width=0.45\textwidth]{imagesMT2014/image_clustering}
  	\caption{(b) Result of K means clustering with data points projected to the top three eigenvectors as major axes. (a) MetaTracts belonging to orientation cluster 1. (c) MetaTracts belonging to orientation cluster 2. MetaTracts in gray (a,c) show context.
  	(d,e) shows the result of distance based clustering on the orientation cluster. (f) shows the combined result. }
  \label{fig:orientation_clustering}
  \end{figure}

@@

\subsection{Choice of clustering techniques}
\label{subsec:clus_choice}
As opposed to the two phase clustering we experimented with using proximity alone and using single step clustering of MetaTracts.
%\begin{figure}[tb]
%\centering
%		\includegraphics[width=0.4\textwidth]{images_pvis/figure7}
%		%\includegraphics[width=0.5\textwidth, trim = 0mm 70mm 0mm 20mm, clip]{imagesMT2014/image7b.pdf}
%	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
%	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
%	\caption{(A) Volume rendering of data set Fig.~\ref{fig:data-char}A.(B) shows a single slice of the data set. (C) shows two of the extracted meshes together. (D) shows the meshes separately.}
%	\label{fig:crop-16-decomp}
%\end{figure}  
%  
Figure~\ref{fig:comparison}a shows the MetaTracts that are hierarchically clustered directly using proximity alone into 10, 15, and 20 clusters (ground truth is 11 clusters) without first performing orientation clustering. As discussed above, single linkage hierarchical clustering in the presence of overlapping fibers tends to create large  incorrect clusters and small(low-cardinality) outliers. When number of clusters ($h$) is 10 two large incorrect clusters are generated and the rest are outliers. As $h$ increases, some appropriate bundles start to form. Even at $h=20$, the top fiber bundles incorrectly cluster together.
%
Figure~\ref{fig:comparison}b shows the MetaTracts, clustered using K-means clustering after being embedded in a $m$-dimensional (lower) space using proximity alone, without a second hierarchical clustering step into 10, 15, and 20 clusters. In the lower dimensional space the spatial context is lost and fibers which are in reality far away are grouped together. Even at $h=20$, no correct bundles are identified.
\begin{figure}[tb]
	\centering
	\includegraphics[width=\linewidth]{images_pvis/comparison_all.pdf}
	\caption{Applying only (a)hierarchical and (b)\added{dimensionality reduction followed by} K-means clustering methods to MetaTracts for various numbers of clusters \added{(distance measure is minimal directed Hausdorffs.}}
	\label{fig:comparison}
\end{figure}

@@

\begin{figure}
\centering
		\includegraphics[width=\linewidth]{images_pvis/figure8.pdf}
		 \vspace{-1.5em}
		%\includegraphics[width=0.5\textwidth, trim = 0mm 70mm 0mm 20mm, clip]{imagesMT2014/image7b.pdf}
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
	\caption{(a) Voxelization of data set 1; (b) a single slice of the data set; (c) two of the extracted meshes together; (d) the meshes rendered separately.}
	\label{fig:crop-16-decomp}
\end{figure}

@@

\begin{figure}[tb]
\centering
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/image8b}
	\includegraphics[width=\linewidth]{images_pvis/dataset2.pdf}
	\caption{Data set with flat thin and compact bundles. (a) shows the volume rendering and a 2D slice with one of the boundaries marked in green, (b) shows the clusters according to individual orientation. (c) shows the complete result. (d) shows the voxelization of (c).}
	\label{fig:prepreg}
\end{figure}

@@

 %glass
We tested our method on three datasets with varying characteristics. We report results on two \replacedWith{due to lack of space}{here and third in the supplemental section}.  
Dataset 1 is described in Sec.~\ref{sec:char_data} (Figure~\ref{fig:data-char}). Dataset 1 has eleven separate fiber bundles (5 along one orientation and 6 along the other). All of them were identified correctly by MetaTracts. With different cross-section sizes and varying degree of curvature of the bundles, dataset 1 is a complicated dataset.
Figure~\ref{fig:orientation_clustering} and Figure~\ref{fig:crop-16-decomp} shows the result of dataset 1 decomposed into two orientation clusters with each orientation cluster further decomposed into $h=10$ clusters followed by voxelization and mesh extraction. \removed{MetaTracts correctly identified the fiber bundles.} Note how the thin and curved purple cluster bundle 7 is extracted well (Figure~\ref{fig:orientation_clustering}e, Figure~\ref{fig:crop-16-decomp}d).

\removed{The hierarchical clustering is robust against over segmentation. For example,}
Figure~\ref{fig:len_dist_crop16} shows the median, minimum and maximum lengths per cluster in orientation 1 (Figure~\ref{fig:orientation_clustering}d) \added{clustered into 10 clusters, the ground truth being  5 clusters}. We generate the correct clusters and outlier bundles with very few elements which can be discarded. As noted in Sec.~\ref{subsec:dist_clustering}, the MetaTracts do not run the entire length of the dataset, the median length of MetaTracts in cluster six (Figure~\ref{fig:len_dist_crop16}) is 200 while the maximum MetaTract length in the bundle is about 500 (measured in units of grid cube length: 2 $\mu m$). Figure~\ref{fig:length_distribution}b shows the length per MetaTract distribution for a particular fiber bundle.

Figure~\ref{fig:prepreg} shows the results on our dataset 2. This dataset has different characteristics, and is dense with \textit{flat and thin} bundles, it consists of 4 bundles along one direction and 5 along the other. The dimensions are 300$\times$350$\times$300 and the datatype is uint16 (dataset 1 is uint8). The bundles are indistinguishable in the original data set Figure~\ref{fig:prepreg}a. A green dotted line shows one of the fiber bundle boundary. Figure~\ref{fig:prepreg}b shows the fiber bundles for the two orientations, Figure~\ref{fig:prepreg}c shows the combined results, and Figure~\ref{fig:prepreg}d shows the volume after the voxelization.

The fiber bundle extraction using MetaTracts was implemented in C++ using ITK. All clustering was done in R. The preprocessing is time consuming and may take several hours. It consists of MetaTract generation and the distance computation. The orientation clustering is robust and done as a part of preprocessing. The hierarchical clustering by itself, on the results of the orientation clusters can be done within an order of a few minutes on Intel Xeon E5-2667 workstation and can be repeated according to user input of $h$ (Sec.~\ref{sec:param_choices}). We store the normalized direction at each fiber point, thus both distance measures can be extracted simultaneously without any extra computations.

>>>>

We demarcate the above queries into two parts: \textit{geometric structure} and \textit{spatial context}. Geometric structure refers to the shape, size, and orientation of a fiber bundle or a group of \replacedWith{bundles}{them}. Spatial context refers to how two or more bundles interact with each other. \replacedWith{For example, spatial context}{It provides} answers \added{to} questions such as: Are these bundles in contact at a particular position in the data? What is the relative orientations of the contacting bundles? 
\replacedWith{Answering}{Providing answers to} the above queries from the volume renderings of XCT datasets or from visual inspection of particular 2D slices is non-trivial, even for experts. We \replacedWith{detail}{present} and evaluate a method which uses visualization techniques to gain insight into our data.
We \replacedWith{advance and interpret}{interpret and advance} techniques from diffusion tensor imaging to extract and visualize geometric structures from \replacedWith{3D X-ray computed tomography}{XCT} data of the woven carbon fiber reinforced composites. The main goal of this work is to expand the state of the art in non-destructive testing through visualization of composite structures in \replacedWith{lower resolutions and thus larger areas}{\textit{complete unit cells}} of woven fabrics.
\begin{figure}[tb]
	\centering
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/algo.png}
	%\includegraphics[width=0.5\textwidth,  trim = 0mm 300mm 0mm 0mm, clip,]{imagesMT2014/image_algo}
	\includegraphics[width=\linewidth]{images_pvis/workflow.pdf}
	\caption{Flow chart of the MetaTracts approach for fiber bundle extraction}
	\label{fig:flowchart}
	\vskip-0.2cm
\end{figure}
Figure~\ref{fig:flowchart} shows an over view of our approach.
\replacedWith
{
Starting from the XCT data, we first generate MetaTracts, cluster the MetaTracts to create fiber bundles and then generate visualizations from the clusters. For a detailed description of the MetaTracts generation process, see Sec.~\ref{subsec:MetaTracts-generation}. For the description of plausible fiber bundle generation see Sec.~\ref{subsec:fiber-bundles}. For a discussion of the visualizations we create for our domain experts see Sec.~\ref{sec:vis}. Sec.~\ref{sec:results} describes our experimental results, Sec.~\ref{sec:param_choices} presents our parameter choices and  Sec.~\ref{sec:user_eval} summarizes our user evaluations.
} 
{
Starting from XCT data, we first generate MetaTracts (see Sec.~\ref{sec:approach}). Then we cluster the MetaTracts to generate and visualize fiber bundles (see Sec.~\ref{subsec:fiber-bundles}). We discuss the visualizations we created for our domain experts in Sec.~\ref{sec:vis}. Sec.~\ref{sec:results} describes our experimental results, Sec.~\ref{sec:param_choices} presents our parameter choices and  Sec.~\ref{sec:user_eval} finally summarizes the user evaluation of our method.
}

@@

\begin{figure}[tb]
\centering
%\includegraphics[width=0.5\textwidth, trim = 0mm 110mm 0mm 0mm, clip,]{images_pvis/figure1}
\includegraphics[width=\linewidth]{images_pvis/data-char.pdf}
\caption
{
\replacedWith
{
Data characteristics: (a) Rendering of the data. (b) 2D slices along the X-axis and (c) 2D slice along the Z-axis. (d) Magnified region marked in green square (c). Multiple fiber bundles cross and are indistinguishable by visual inspection alone. 
}
{
Data characteristics: (a) Rendering of dataset D1. (b, c) 2D slices along Z- and X-axis. (d) Zoom in of the green region marked in (c). Multiple fiber bundles cross and are indistinguishable by visual inspection alone. 
}
}
\label{fig:data-char}
\vskip-0.2cm
\end{figure}

\section {Data Characteristic and Assumptions}
\label{sec:char_data}
\replacedWith
{
Figure \ref{fig:data-char} shows \replacedWith{dataset 1}{D1} with woven fiber bundles. \replacedWith{Dataset 1}{D1} is 450$\times$300$\times$500 voxels in size with isotropic resolution of 2$\mu m$. The data type is unsigned 8 bit integer. Figure~\ref{fig:data-char} clearly shows the weaving structure of the composite unit cell as regional subset of the recurring fiber bundle weaving pattern. The unit cell is used as a basis structure for CFRP manufacturing. The alignment and weaving pattern of the fiber bundles in the cloth influences the strength and stiffness properties of the resulting material. Figure~\ref{fig:data-char}a shows a volume rendering of the dataset. Figure~\ref{fig:data-char}b shows 2D slices along the $X$- and Figure~\ref{fig:data-char}c in $Z$-axis. Figure~\ref{fig:data-char}d shows the magnified version of the green region of interest. 
The resolution of \replacedWith{dataset 1}{D1} renders the individual carbon fibers in a fiber bundle hard to resolve in the 2D slice images. Figure~\ref{fig:data-char}d contains two bundles going in opposite directions. The separation between two fiber bundles is barely visible. The carbon fibers themselves do not differ much from the underlying epoxy matrix in terms of attenuation which poses an additional challenge for characterization. The fiber bundles may differ in terms of the amount of fibers in the bundle. Figure~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern, the fiber bundles cross each other in different orientations. In general, the number of orientations is limited by the weaving pattern. Nevertheless, the weaving pattern may cause the individual bundles to be curved. In consequence individual fibers may be adjacent in Euclidean distance but belong to different bundles. 
We make the following assumptions on our data:
}
{
Figure~\ref{fig:data-char} shows dataset D1 with woven fiber bundles. The size of D1 is 450$\times$300$\times$500 voxels with isotropic resolution of 2 $\mu m$ and 8 bit unsigned integer scalars. Figure~\ref{fig:data-char} clearly shows the recurring fiber bundle weaving pattern of the composite unit cell used for manufacturing. Figure~\ref{fig:data-char}a shows a volume rendering of the dataset. Figure~\ref{fig:data-char}b and ~\ref{fig:data-char}c show 2D slices along the $X$- and the $Z$-axis respectively. Figure~\ref{fig:data-char}d zooms into the green region of interest. 
The resolution of D1 renders the individual carbon fibers in a fiber bundle hard to resolve in the 2D slice images. Figure~\ref{fig:data-char}d contains two bundles going in opposite directions. Also the separation between two fiber bundles is barely visible. The fiber bundles may differ in terms of the amount of fibers in the bundle. Figure~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern, the fiber bundles cross each other in different orientations. The number of orientations is defined by the weaving pattern and typically consists of two main orientations. The weaving pattern may cause individual bundles to be curved. In consequence, individual fibers may be adjacent in Euclidean distance but belong to different bundles. 
We make the following \textit{assumptions} on our data:
}

@@

\begin{figure}[tb]
\centering
%\includegraphics[width=0.45\textwidth]{images_pvis/figure3}
%\includegraphics[width=0.5\textwidth,  trim = 0mm 90mm 0mm 0mm, clip]{images_pvis/figure3}
\includegraphics[width=\linewidth]{images_pvis/reliable_hessian.pdf}
 \vspace{-1.5em}
\caption{Reliable Hessians. (a) Colored according to Orientation vector mapped to RGB. (b\added{,c}) 2D slice\added{s} along Z\replacedWith{ and (c) along}{-} and X axis. (d) \replacedWith{Magnified}{Zoom in of the green} region marked in \replacedWith{green square}{(c)}.}
\label{fig:reliable_hessian}
\vskip-0.2cm
\end{figure}

@@

\subsection {MetaTracts generation}
\label{subsec:MetaTracts-generation}
In this section we explain the process of MetaTracts generation in $\mathbb{R}^2$ (the procedure trivially extends to $\mathbb{R}^3$). In $\mathbb{R}^2$, all the above properties hold except that cylinders are replaced by rectangles. The reliable Hessian step generates a local orientation vector in $\mathbb{R}^2$ and a reliability measure $R_H$[0,1]. In Figure~\ref{fig:algo}, regions with unreliable Hessian have been marked in blue. Figure~\ref{fig:algo}c shows \replacedWith{a single}{an individual} cylinder of a MetaTract. Let the seed point associated with the MetaTract be grid point $C_{p}$ (property 2 in Sec.~\ref{subsec:mtprop}). The local orientation at $C_p$ as computed in the step above is $N_P$ and is given by the dark green arrow. We generate a rectangle of length $L$ and radius $R$ (properties 3 and 5 in Section~\ref{subsec:mtprop}). 
%
\begin{figure}[tb]
  \begin{center}
 \includegraphics[ width=\linewidth]{images_pvis/mt_gen.pdf}
  \end{center}
  \vspace{-1em}
 \caption{MetaTracts generation: (b) \replacedWith{MetaTract generation, region with $R_H$ less than threshold}{Generation process}. (a) \replacedWith{Close up of}{Zoom in of the white} marked region in (b). (c) \replacedWith{Parts of a single MetaTract}{Individual cylinder}.}
  \label{fig:algo}
    %\vspace{-1.5em}
    \vskip-0.2cm
\end{figure}
The set of vertices which are in the green region but not overlapped by blue are possible \textit{``candidate vertices"} for the start point of the next cylinder in a particular MetaTract. From these we select another grid vertex which will be the start point for the next cylinder.
%\begin{figure}
%\centering
% \includegraphics[width=0.25\textwidth]{imagesMetaTracts/angle-diff.png}
%\caption{Angle between the local orientation of two cylinders that are part of the same MetaTract.}
%\label{fig:angle-diff}
%\end{figure}
Based on assumptions on our data (local orientation and connectivity), the order of the candidate vertices is based on the following \textit{characteristics}:

@@

\begin{figure}[tb]
\centering
	\includegraphics[width=\linewidth]{images_pvis/figure5.pdf}
	%\includegraphics[width=0.45\textwidth,  trim = 0mm 100mm 0mm 50mm, clip]{imagesMT2014/image_length}
	\caption{(a) MetaTracts color-coded according to their mean orientations. (b) length distribution of individual MetaTracts for a particular bundle (unit for length is grid cube edge length:  $2\mu m$).}
	\label{fig:length_distribution}
	\vskip-0.2cm
\end{figure}

@@

\begin{figure}[tb] 
  \centering  	
  \includegraphics[width=\linewidth]{images_pvis/clustering.pdf}
   \vspace{-1.5em}
  %\includegraphics[width=0.45\textwidth]{imagesMT2014/image_clustering}
  	\caption{(b) Result of K means clustering with data points projected to the top three eigenvectors as major axes. (a) MetaTracts belonging to orientation cluster 1. (c) MetaTracts belonging to orientation cluster 2. MetaTracts in gray (a,c) show context.
  	(d,e) shows the result of distance based clustering on the orientation cluster. (f) shows the combined result. }
  \label{fig:orientation_clustering}
  \vskip-0.2cm
  \end{figure}

@@

\subsection{Choice of clustering techniques}
\label{subsec:clus_choice}
As opposed to the two phase clustering we experimented with using proximity alone and using single step clustering of MetaTracts.
%\begin{figure}[tb]
%\centering
%		\includegraphics[width=0.4\textwidth]{images_pvis/figure7}
%		%\includegraphics[width=0.5\textwidth, trim = 0mm 70mm 0mm 20mm, clip]{imagesMT2014/image7b.pdf}
%	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
%	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
%	\caption{(A) Volume rendering of data set Fig.~\ref{fig:data-char}A.(B) shows a single slice of the data set. (C) shows two of the extracted meshes together. (D) shows the meshes separately.}
%	\label{fig:crop-16-decomp}
%\end{figure}  
%  
Figure~\ref{fig:comparison}a shows the MetaTracts that are hierarchically clustered directly using proximity alone into 10, 15, and 20 clusters (ground truth is 11 clusters) without first performing orientation clustering. As discussed above, single linkage hierarchical clustering in the presence of overlapping fibers tends to create large  incorrect clusters and small(low-cardinality) outliers. When number of clusters ($h$) is 10 two large incorrect clusters are generated and the rest are outliers. As $h$ increases, some appropriate bundles start to form. Even at $h=20$, the top fiber bundles incorrectly cluster together.
%
Figure~\ref{fig:comparison}b shows the MetaTracts, clustered using K-means clustering after being embedded in a $m$-dimensional (lower) space using proximity alone, without a second hierarchical clustering step into 10, 15, and 20 clusters. In the lower dimensional space the spatial context is lost and fibers which are in reality far away are grouped together. Even at $h=20$, no correct bundles are identified.
\begin{figure}[tb]
	\centering
	\includegraphics[width=\linewidth]{images_pvis/comparison_all.pdf}
	\caption{Applying only (a)hierarchical and (b)\added{dimensionality reduction followed by} K-means clustering methods to MetaTracts for various numbers of clusters \added{(distance measure is minimal directed Hausdorffs.}}
	\label{fig:comparison}
	\vskip-0.2cm
\end{figure}

@@

\begin{figure}
\centering
		\includegraphics[width=\linewidth]{images_pvis/figure8.pdf}
		 \vspace{-1.5em}
		%\includegraphics[width=0.5\textwidth, trim = 0mm 70mm 0mm 20mm, clip]{imagesMT2014/image7b.pdf}
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
	\caption{(a) Voxelization of data set 1; (b) a single slice of the data set; (c) two of the extracted meshes together; (d) the meshes rendered separately.}
	\label{fig:crop-16-decomp}
	\vskip-0.2cm
\end{figure}

@@

\begin{figure}[tb]
\centering
	%\includegraphics[width=0.45\textwidth]{imagesMT2014/image8b}
	\includegraphics[width=\linewidth]{images_pvis/dataset2.pdf}
	\caption{Data set with flat thin and compact bundles. (a) shows the volume rendering and a 2D slice with one of the boundaries marked in green, (b) shows the clusters according to individual orientation. (c) shows the complete result. (d) shows the voxelization of (c).}
	\label{fig:prepreg}
	\vskip-0.2cm
\end{figure}

@@

 %glass
We tested our method on three datasets with varying characteristics. We report results on two \replacedWith{due to lack of space}{here and third in the supplemental section}.  
Dataset \added{D}1 is described in Sec.~\ref{sec:char_data} (Figure~\ref{fig:data-char}). \replacedWith{Dataset 1}{D1} has eleven separate fiber bundles (5 along one orientation and 6 along the other). All of them were identified correctly by MetaTracts. With different cross-section sizes and varying degree of curvature of the bundles, \replacedWith{dataset 1}{D1} is a complicated dataset.
Figure~\ref{fig:orientation_clustering} and Figure~\ref{fig:crop-16-decomp} shows the result of \replacedWith{dataset 1}{D1} decomposed into two orientation clusters with each orientation cluster further decomposed into $h=10$ clusters followed by voxelization and mesh extraction. \removed{MetaTracts correctly identified the fiber bundles.} Note how the thin and curved purple cluster bundle 7 is extracted well (Figure~\ref{fig:orientation_clustering}e, Figure~\ref{fig:crop-16-decomp}d).

\removed{The hierarchical clustering is robust against over segmentation. For example,}
Figure~\ref{fig:len_dist_crop16} shows the median, minimum and maximum lengths per cluster in orientation 1 (Figure~\ref{fig:orientation_clustering}d) \added{clustered into 10 clusters, the ground truth being  5 clusters}. We generate the correct clusters and outlier bundles with very few elements which can be discarded. As noted in Sec.~\ref{subsec:dist_clustering}, the MetaTracts do not run the entire length of the dataset, the median length of MetaTracts in cluster six (Figure~\ref{fig:len_dist_crop16}) is 200 while the maximum MetaTract length in the bundle is about 500 (measured in units of grid cube length: 2 $\mu m$). Figure~\ref{fig:length_distribution}b shows the length per MetaTract distribution for a particular fiber bundle. 
Figure~\ref{fig:prepreg} shows the results on our dataset 2 \added{(D2)}. This dataset has different characteristics, and is dense with \textit{flat and thin} bundles, it consists of 4 bundles along one direction and 5 along the other. The dimensions are 300$\times$350$\times$300 and the datatype is uint16 (\replacedWith{dataset 1}{D1} is uint8). The bundles are indistinguishable in the original data set Figure~\ref{fig:prepreg}a. A green dotted line shows one of the fiber bundle boundary. Figure~\ref{fig:prepreg}b shows the fiber bundles for the two orientations, Figure~\ref{fig:prepreg}c shows the combined results, and Figure~\ref{fig:prepreg}d shows the volume after the voxelization.

\removed{The fiber bundle extraction using }MetaTracts \replacedWith{was}{are} implemented in C++ using ITK. \removed{All }Clustering \replacedWith{was}{is} done in R. The preprocessing is time consuming and may take several hours. It consists of MetaTract generation and the distance computation. The orientation clustering is robust and done as a part of preprocessing. The hierarchical clustering by itself, on the results of the orientation clusters can be done within an order of a few minutes on Intel Xeon E5-2667 workstation and can be repeated according to user input of $h$ (Sec.~\ref{sec:param_choices}). We store the normalized direction at each fiber point, thus both distance measures can be extracted simultaneously without any extra computations.
